Subjective Evaluations of Financial Satisfaction and Well-being: A Study of the US General Social Survey

### 2023-03-10 outputs/paper/paper3.qmd
Complete first pass at social status. Fixed up minor errors

| Variable | New Name               | Description                                                                             | Example               |
|------------------|------------------|-------------------|------------------|
| satfin   | financial_satisfaction | Respondent's satisfaction with their current financial situation                        | Pretty Well Satisfied |
| finalter | financial_change       | Changes in respondent's financial situation                                             | Stayed the same       |
| finrela  | financial_compare      | Comparing respondent's family income with American families in general                  | Above Average         |
| class    | social_class           | self identified social class the respondent belong in                                   | Middle Class          |
| class1   | social_class1          | self identified social class the respondent belong in (more granular than social_class) | Upper Middle Class    |

@@

First, we are interested in the correlation between family income and social class. Looking at the 4 band classification of social class, we see in @fig-socialclass-1 that less than 10% of respondents classified themselves as **lower class**. Digging deeper into the family income of this segment in @fig-income-social , the majority of respondents (\~90%) in this class are making less than \$40K annually. There are 37% of respondents who classified themselves as **working class**. There is a wide range of their family income, with the majority (\~71%) clustered between \$10K to \$75K. Almost half of the respondent have classified themselves as **middle class**, with family income clustered between \$60K to \$170K (\~57%). It is interesting to note that those making over \$170K have the highest percentage (15%) compared across income bands for those who ranked themselves in this class. Lastly, out of the 4% of the respondents who responded that they belong to the **upper class**, most of them have family income over \$170K (67%). An interesting observation in this social class is that more than 10% of respondents making less than \$40K a year also put themselves as upper class.

The 6 band social class survey expands the middle class into three categories: lower middle class, middle class, and upper middle class. Comparing the distribution between the two social class variables (@fig-socialclass), there are movements across the bands.

```{r}
#| message: false
#| echo: false
#| warning: false

survey_data_2021 |>
  filter(is.na(social_class1)) |>
  group_by(social_class) |>
  count() |>
  ungroup() |>
  mutate(Percentage = n / sum(n))

survey_data_2021 |>
  filter(!is.na(social_class), !is.na(income)) |>
  group_by(social_class, income) |>
  count() |>
  ungroup() |>
  group_by(social_class) |>
  mutate(Percentage = n / sum(n)) |>
  mutate(Cum = cumsum(Percentage)) |>
  filter (social_class == "Upper Class")
```

@@

@@

Social class over the years

```{r}
#| message: false
#| echo: false
#| warning: false

combined_data |>
  select(year, social_class) |>
  group_by(year) |>
  count(social_class) |>
  filter(!is.na(social_class)) |>
  mutate(percent = n / sum(n)) |>
  ggplot(aes(x = year, y = percent, color = social_class)) +
  geom_smooth(method=lm, se=FALSE) + 
  geom_line() +
  scale_y_continuous(labels=scales::percent) +
  labs(
    xlabel = "Year",
    ylabel = "Percentage of Responses",
    color = "Social Class"
  )

```

```{r}
#| message: false
#| echo: false
#| warning: false

survey_data_2021 |>
  filter(!is.na(income), !is.na(social_class)) |>
  ggplot(aes(x = social_class, group = income)) +
  geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat="count") + 
  scale_y_continuous(labels=scales::percent) +
  ylab("Percentage of response") +
  xlab("Financial Satisfaction") +
  theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  theme(legend.position="none") +
  facet_grid(~income)
```

```{r}
#| message: false
#| echo: false
#| warning: false

# survey_data_2021 |>
#   filter(!is.na(social_class)) |>
#   ggplot(aes(x = social_class, y = realinc)) +
#   geom_boxplot() +
#   geom_jitter(alpha = 0.3, width = 0.10, height = 0)
#   #geom_point(alpha = 0.2) +
  #geom_violin()
```

@@

### (Economic) The Rich are getting Richer, and the Poor are getting Poorer

@@

### (Psychological) Fleishman Effect

@@

## Limitations and Future Direction

@@

We propose adding the questions "If you went to college or university, what was your major?" "Do you regret your major? (Y/N/Don't know)" to the survey. Adding this question to our study would provide more detailed insight on reference groups, since we could see how people who enrolled in each major compare themselves to others, and, with more information on family history, we could also gather insight on which fields individuals from different social classes pursue in higher education. We could also see how people across different income ranges and classes experience regret about their decisions.

Additional financial variables in survey for analysis

Connect with US census data to check representativeness

Connect with other macroeconomic indicators - e.g. inflation, stock markets

>>>>

| Variable | New Name               | Description                                                                             | Example               |
|------|---------------|----------------------|---------------|
| satfin   | financial_satisfaction | Respondent's satisfaction with their current financial situation                        | Pretty Well Satisfied |
| finalter | financial_change       | Changes in respondent's financial situation                                             | Stayed the same       |
| finrela  | financial_compare      | Comparing respondent's family income with American families in general                  | Above Average         |
| class    | social_class           | self identified social class the respondent belong in                                   | Middle Class          |
| class1   | social_class1          | self identified social class the respondent belong in (more granular than social_class) | Upper Middle Class    |

@@

First, we are interested in the correlation between family income and social class. Looking at the 4 band classification of social class, we see in @fig-socialclass-1 that less than 10% of respondents classified themselves as **lower class**. Digging deeper into the family income of this segment in @fig-income-social, the majority of respondents (\~90%) in this class are making less than \$40K annually. There are 37% of respondents who classified themselves as **working class**. There is a wide range of their family income, with the majority (\~71%) clustered between \$10K to \$75K. Almost half of the respondent have classified themselves as **middle class**, with family income clustered between \$60K to \$170K (\~57%). It is interesting to note that those making over \$170K have the highest percentage (15%) compared across income bands for those who ranked themselves in this class. Lastly, out of the 4% of the respondents who responded that they belong to the **upper class**, most of them have family income over \$170K (67%). An interesting observation in this social class is that more than 10% of respondents making less than \$40K a year also put themselves as upper class.

The 6 band social class survey expands the middle class into three categories: lower middle class, middle class, and upper middle class. Comparing the distribution between the two social class variables (@fig-socialclass), there are some movements across the bands; for example, there are less respondents who rated themselves as lower class and upper class. Looking at the income levels associated with **upper class** in this 6-band classifcation, there are some peculiar observations. While most individuals in this social class responded with making over \$170K annually, 20% of them reported family income less than \$10K.

@@

Looking at the correlation between social class and financial satisfaction, we saw the expected trend in @fig-socialclass-financial-1 that as individuals move up to a higher social class, they report higher financial satisfaction with their current situation. However, when we look at the detailed social class breakdown in @fig-socialclass-financial-2, we noticed a trend reversal for **upper class**, where close to 20% of the respondents reported not satisfied at all with their financial situation, which is much higher compared to upper middle class with only less than 2% dissatisfaction. We will hypothesize potential explanations for this observation in the the discussion section below on [The Fleishman Effect].

@@

Looking at the percentages of respondents in different social class over the years in @fig-socialclass-overtime, we discovered a declining trend for working class and middle class, while the percentage of those who reported as lower class has been increasing over time. Upper class percentage stays mainly constant from 1972 to 2021 at around 3-4%.

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: fig-socialclass-overtime
#| fig-cap: "Percentage of respondents in each social class (4-bands) over time"

generate_time_series_graph(
  ggplot_data =
    combined_data |>
    select(year, social_class) |>
    group_by(year, social_class) |>
    count() |>
    filter(!is.na(social_class)) |>
    group_by(year) |>
    mutate(percent = n / sum(n)) |>
    ggplot(aes(x = year, y = percent, color = social_class)),
  xlabel = "Year",
  ylabel = "Percentage of Responses",
  color_label = "Social Class",
  position = NULL,
  nrow = NULL
)
```

@@

## Marginalization

-   Men are doing better than women

-   Level of education is a big factor in determining financial satisfaction

-   Generation gap?

-   Even though GDP rising, we don't see significant improvement in financial satisfaction over time

-   Increase of lower class % over time

## (Economic) Polarization of Financial Satisfaction

@@

-   @fig-overtime polarization of financial situation - people are either getting better or getting worse, but not many of them are staying "average". Dissatisfaction overall is rising.

-   Not much changes in social class across the years

-   Those who are making lower income are less satisfied

## (Psychological) Fleishman Effect

@@

## Future Research Directions

@@

We propose adding the questions "If you went to college or university, what was your major?" "Do you regret your major? (Y/N/Don't know)" to the survey. Adding this question to our study would provide more detailed insight on reference groups, since we could see how people who enrolled in each major compare themselves to others, and, with more information on family history, we could also gather insight on which fields individuals from different social classes pursue in higher education. We could also see how people across different income ranges and classes experience regret about their decisions. Please see [Appendix] to view the full supplementary survey that we have put together.

-   Additional financial variables in survey for analysis

-   Connect with US census data to check representativeness

-   Connect with other macroeconomic indicators - e.g. inflation, stock markets
